Multiresolution Neural Field Estimation

### 2011-02-28 Multiresolution_framework/ltx/manuscript.tex
added cyan for cutting texts
more edits to derivation

\usepackage{color}
\newcommand{\mike}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}} 
\newcommand{\dean}[1]{\textsf{\emph{\textbf{\textcolor{green}{#1}}}}} 
\newcommand{\parham}[1]{\textsf{\emph{\textbf{\textcolor{blue}{#1}}}}} 
\newcommand{\ken}[1]{\textsf{\emph{\textbf{\textcolor{magenta}{#1}}}}} 
\newcommand{\cut}[1]{\textsf{\emph{\textbf{\textcolor{cyan}{#1}}}}}

@@

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Neural field model, Expectation Maximization (EM) algorithm, Multiresolution Approximation (MRA), wavelets.
\end{IEEEkeywords}

@@

\section{Introduction}
\IEEEPARstart{T}{his} paper introduces a multiresolution data-driven framework for neural field modeling. Neural field models describe the meso and macroscopic dynamics of the human brain. Modeling the brain at this level is extremely important, since diseases such as epilepsy and \parham{Parkinson's disease} are manifested at this scale through pathological oscillations. The ability to create patient-specific neural field models will not only contribute to our knowledge base of poorly understood diseases such as epilepsy, but may also enable the development of new treatment strategies. This is particularly poignant with the advent of new devices that use therapeutic electrical stimulation. Presently, stimulation strategies for devices using corrective or modulating electrical stimulation operate in an open-loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists an enormous potential to improve the performance of these devices using systems theory. The design of a stimulation strategy using systems theory requires a model. Neural mass and neural field models are the ideal candidates for estimation and control algorithms. It is expected that parameters of the neural fields models will be patient specific and will therefore need to be inferred from data.

@@

\section{Integro-Difference Equation Neural Field Model}
The stochastic integro-difference equation (IDE) form of the Amari neural field  formulation~\cite{Amari1977} is given by (see \dean{cite NeuroImage paper} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Synaptic dynamics are included through the parameter $\xi=\left(1-\frac{ T_s}{\tau}\right)$, where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function $f\left(.\right)$. \dean{The activation function is of the form $f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}) + v_0$}. The term $e_t(\mathbf r)$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$, where $\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$  denotes a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$~\cite{Rasmussen2005}. For simplicity, \cut{we denote the current time step by $t$ and the future time step by $t+1$.} The observation equation describing the electrophysiological recordings is given by 
\begin{equation}\label{eq:ObservationEquation}
	\mathbf y_t = \int_{\Omega} { m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, d\mathbf{r}'} + \boldsymbol\epsilon_t, 
\end{equation}
where $\mathbf{y}_{t} = [y_t(\mathbf{r}_1) y_t(\mathbf{r}_2) \cdots y_t(\mathbf{r}_{n_y})]^\top$, compiled at $n_{y}$ spatial locations at time $t$ via the observation kernel $m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right)$, is corrupted by an i.i.d normally distributed zero-mean white noise $\boldsymbol{\epsilon}_{t}\sim \mathcal{N}\left(\mathbf{0},\mathbf{\Sigma}_{\epsilon}\right)$ with $\mathbf{\Sigma}_{\epsilon}=\sigma_{\epsilon}^2\mathbf I_{n_y} $. The superscript $\top$ denotes the transpose operator.

@@

\section{State and Parameter Estimation}
A state-space representation of the IDE model allows the use of the Expectation Maximisation (EM) algorithm \cite{Dempster1977,Shumway2000} to infer both the kernel and the field from the electrophysiological data. The EM algorithm is a two part iterative algorithm that yields the maximum likelihood estimate of the connectivity kernel parameters and the posterior distribution of the field weights. In the M-step a lower bound $\mathcal{Q}\left(\boldsymbol\theta,\boldsymbol\theta'\right)$ of the complete-data log-likelihood  $p\left(\mathbf{X},\mathbf{Y} \mid \boldsymbol\theta \right)$, is maximised under the assumption of the system parameters being $\boldsymbol\theta'$. In the E-step, given an estimate of $\boldsymbol\theta$, the Rauch Tung Striebel (RTS) smoother \cite{Gibsona2005} can be employed to determine the posterior state sequence. 
The state-space representation of the IDE defined by \eqref{eq:StateEquation} and \eqref{eq:ReducedObservationEquation} has a quadratic $\mathcal Q\left(\boldsymbol\theta,\boldsymbol \theta'\right)$ in a form of
\begin{equation}
 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta
\end{equation}
with a maximum at
\begin{align}
 \boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0^\top-\boldsymbol\upsilon_1^\top).
\end{align}
Where
\begin{align}\label{eq:upsilon0}
 \boldsymbol\upsilon_0&=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\\ 
 \boldsymbol\upsilon_1&=\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\label{eq:upsilon1} \\
 \boldsymbol\Upsilon&=T_s\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\mathbf{U}^{\mathsf T} \boldsymbol\Lambda_{x}^{-1}\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf{U}]_{j,i}\label{eq:Upsilon}
\end{align}
and $\beta$ is constant with respect to the connectivity kernel parameters. The $\left(i,j\right)^{th}$ element of the  block matrix $\mathbf U_{n_x \times n_x}$ is given by
\begin{align}
\left[ \mathbf U\right] _{i,j}&=\int_{\boldsymbol \Omega}[\mu(\mathbf r)]_i[\int_{\boldsymbol\Omega}\boldsymbol\mu\left(\mathbf{r'}\right)\boldsymbol\lambda^\top\left(\mathbf{r-r'}\right) d\mathbf{r'}]_{j:}d\mathbf r.
\end{align}
where $[.]_{j:} $ denotes $j^{th}$ row of the matrix. The second terms in equations (\ref{eq:upsilon0}-\ref{eq:Upsilon}) can be computed before EM implementation which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}. $\Xi$-variables can be calculated using the smoother outputs, i.e., state estimates, $\hat{\mathbf x}_t$ along with their covariance matrix, $\mathbf P_t$ and the cross-covariance matrix $\mathbf M_t$ using \cite{Gibsona2005}
\begin{align}\label{eq:Xivariables}
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right)
\end{align}
\section{Simulation}
This section demonstrates the identification of the proposed IDE model. The MRA of a synthetic model is demonstrated, which is able to represent the system at different spatial scales. The spectral analysis was used to choose the spatial scale adequate to represent the spatial field.

>>>>

\usepackage{color}
\newcommand{\mike}[1]{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}} 
\newcommand{\dean}[1]{\textsf{\emph{\textbf{\textcolor{green}{#1}}}}} 
\newcommand{\parham}[1]{\textsf{\emph{\textbf{\textcolor{blue}{#1}}}}} 
\newcommand{\ken}[1]{\textsf{\emph{\textbf{\textcolor{magenta}{#1}}}}}

@@

% Note that keywords are not normally used for peerreview papers.
\begin{IEEEkeywords}
Neural field model, multiresolution approximation (MRA), expectation maximization (EM) algorithm, wavelets.
\end{IEEEkeywords}

@@

\section{Introduction}
\IEEEPARstart{T}{his} paper introduces a multiresolution data-driven framework for neural field modeling. Neural field models describe the meso and macroscopic dynamics of the human brain. Modeling the brain this level is extremely important, since diseases such as epilepsy and Parkinson's disease are manifested at this scale through pathological oscillations. The ability to create patient-specific neural field models will not only contribute to our knowledge base of poorly understood diseases such as epilepsy, but may also enable the development of new treatment strategies. This is particularly poignant with the advent of new devices that use therapeutic electrical stimulation. Presently, stimulation strategies for devices using corrective or modulating electrical stimulation operate in an open-loop, where stimulation parameters are chosen using a process of trial and error. Therefore, there exists an enormous potential to improve the performance of these devices using systems theory. The design of a stimulation strategy using systems theory requires a model. Neural mass and neural field models are the ideal candidates for estimation and control algorithms. It is expected that parameters of the neural fields models will be patient specific and will therefore need to be inferred from data.

@@

\section{Integro-Difference Equation Neural Field Model}
The stochastic integro-difference equation (IDE) form of the Amari neural field  formulation~\cite{Amari1977} is given by (see \dean{cite NeuroImage paper} for a full derivation)
\begin{equation}\label{eq:DiscreteTimeModel}
	v_{t+T_s}\left(\mathbf{r}\right) = 
	\xi v_t\left(\mathbf{r}\right) + 
	T_s \int_\Omega { 
	    w\left(\mathbf{r},\mathbf{r'}\right)
	    f\left(v_t\left(\mathbf{r}'\right)\right) 
	\, d\mathbf{r}'} 
	+ e_t\left(\mathbf{r}\right), 
\end{equation}
where the post-synaptic membrane voltage at time $t$ of a population of neurons at position $\mathbf r$ is denoted $v_t\left(\mathbf r\right)$. Synaptic dynamics are included through the parameter $\xi=\left(1-\frac{ T_s}{\tau}\right)$, where $\tau$ is the synaptic time constant and $T_s$ is the sampling time. The connectivity strength between neural populations at a distance $\mid\mathbf{r}-\mathbf{r'}\mid$ is described by the kernel $w\left(\mathbf{r}-\mathbf{r}'\right)$. The connectivity kernel is taken as a ``Mexican hat'' function, which describes local excitation and lateral inhibition \cite{Amari1977}. The firing rate of the presynaptic neurons is related to the post-synaptic membrane potential by the activation function $f\left(.\right)$. \dean{The activation function is of the form $f(v_t(\mathbf{r})) = \varsigma v_t(\mathbf{r}) + v_0$}. The term $e_t(\mathbf r)$ is an $i.i.d.$ disturbance such that $e_t(\mathbf{s})\sim\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$, where $\mathcal{GP}(\mathbf 0,\eta(\mathbf{r}-\mathbf{r'}))$  denotes a zero mean spatial Gaussian process with covariance function $\eta(\mathbf{r}-\mathbf{r'})$~\cite{Rasmussen2005}. For simplicity, we denote the current time step by $t$ and the future time step by $t+1$. The observation equation describing the electrophysiological recordings is given by 
\begin{equation}\label{eq:ObservationEquation}
	\mathbf y_t = \int_{\Omega} { m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right) v_t\left(\mathbf{r}'\right) \, d\mathbf{r}'} + \boldsymbol\epsilon_t, 
\end{equation}
where $\mathbf{y}_{t} = [y_t(\mathbf{r}_1) y_t(\mathbf{r}_2) \cdots y_t(\mathbf{r}_{n_y})]^\top$, compiled at $n_{y}$ spatial locations at time $t$ via the observation kernel $m\left(\mathbf{r}_{n_y}-\mathbf{r}'\right)$, is corrupted by an i.i.d normally distributed zero-mean white noise $\boldsymbol{\epsilon}_{t}\sim \mathcal{N}\left(\mathbf{0},\mathbf{\Sigma}_{\epsilon}\right)$ with $\mathbf{\Sigma}_{\epsilon}=\sigma_{\epsilon}^2\mathbf I_{n_y} $. The superscript $\top$ denotes the transpose operator.

@@

\section{State and Parameter Estimation}
The state-space representation allows the use of the expectation maximization (EM) algorithm \cite{Dempster1977,Shumway2000} to infer both the kernel and the field from the electrophysiological data. The EM algorithm is a two-part iterative algorithm that yields the maximum likelihood estimate of the connectivity kernel parameters (M-step) and the posterior distribution of the field weights (E-step). The M-step finds an estimate of $\boldsymbol\theta$ that maximizes the lower bound, $\mathcal{Q}\left(\boldsymbol\theta,\boldsymbol\theta'\right)$, of the complete-data log-likelihood,  $p\left(\mathbf{X},\mathbf{Y} \mid \boldsymbol\theta \right)$. The E-step uses the estimate of $\boldsymbol\theta$, $\boldsymbol\theta'$ and the Rauch Tung Striebel (RTS) smoother \cite{Gibsona2005} to estimate the posterior state sequence. 
The bound on the log-likelihood is a quadratic in the form of
\begin{equation}
 \mathcal Q\left(\boldsymbol \theta,\boldsymbol\theta'\right)=\beta+2T_s\left(\boldsymbol\upsilon_0-\boldsymbol\upsilon_1\right)\boldsymbol\theta-T_s\boldsymbol\theta^\top\boldsymbol\Upsilon\boldsymbol\theta,
\end{equation}
with a maximum at
\begin{align}
 \boldsymbol \theta= \boldsymbol\Upsilon^{-\top}(\boldsymbol\upsilon_0^\top-\boldsymbol\upsilon_1^\top),
\end{align}
where
\begin{align}\label{eq:upsilon0}
 \boldsymbol\upsilon_0&=\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_0]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\\ 
 \boldsymbol\upsilon_1&=\xi\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf U]_{j,i}\label{eq:upsilon1} \\
 \boldsymbol\Upsilon&=T_s\sum_{i,j=1}^{n_x}[\boldsymbol\Xi_1]_{i,j}[\mathbf{U}^{\mathsf T} \boldsymbol\Lambda_{x}^{-1}\Sigma_{w}^{-1}\boldsymbol\Lambda_{x}^{-1}\mathbf{U}]_{j,i}\label{eq:Upsilon}.
\end{align}
The $\left(i,j\right)^{th}$ element of the block matrix $\mathbf U_{n_x \times n_x}$ is given by
\begin{align}
\left[ \mathbf U\right] _{i,j}&=\int_{\boldsymbol \Omega}\left[\mu(r) \right]_i \left[\int_{\boldsymbol\Omega} \boldsymbol\mu\left(r'\right)\boldsymbol \lambda^\top \left(\mathbf{r-r'}\right) dr'\right]_{j:} dr.
\end{align}
where $[.]_{j:} $ denotes $j^{th}$ row of the matrix. The second terms in equations (\ref{eq:upsilon0}-\ref{eq:Upsilon}) can be computed before EM implementation which increases the speed of the M-step significantly compared to the method in \cite{Dewar2009}. $\Xi$-variables can be calculated using the smoother outputs, i.e., state estimates, $\hat{\mathbf x}_t$ along with their covariance matrix, $\mathbf P_t$ and the cross-covariance matrix $\mathbf M_t$ using \cite{Gibsona2005}
\begin{align}\label{eq:Xivariables}
\boldsymbol\Xi_0&=\sum_{t=0}^{T-1}\left(\mathbf M_{t+1}+\mathbf{\hat x}_t\mathbf{\hat x}_{t+1}^\top\right) \\
 \boldsymbol\Xi_1&=\sum_{t=0}^{T-1}\left(\mathbf P_t+\mathbf{\hat x}_t\mathbf{\hat x}_t^\top\right)
\end{align}
\section{Simulation}
This section demonstrates the identification of the proposed IDE model. The MRA of a synthetic model is demonstrated, which is able to represent the system at different spatial scales. The spectral analysis was used to choose the spatial scale adequate to represent the spatial field.
